---
title: "Cell and cell-type level scores and p-values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cell and cell-type level scores and p-values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

SCADS produces two kinds of inference. At the level of a single cell, a disease
relevance score with a p-value. At the level of a cell type, a test of whether
the cells within it differ from one another in disease relevance. This vignette
gives both formulas and shows how to run them.

## The cell score

Each cell's score is the enrichment of disease heritability in its open regions
relative to genome-wide, reconstructed from the topic model:

$$c_i = \frac{\sum_k\sum_j l_{ik} I_{kj} e_k v_j}{\sum_k\sum_j l_{ik} I_{kj} v_j}
      = \sum_k w_{ik}\, e_k, \qquad \sum_k w_{ik} = 1$$

where $l_{ik}$ is the cell's loading on topic $k$, $I_{kj}$ marks region $j$ as
open in topic $k$, and $e_k$ is that topic's heritability enrichment from S-LDSC.
The weights $w_{ik} \propto l_{ik} s_k$ depend on the loadings and the topic
annotation sizes $s_k$, and sum to one, so $c_i = 1$ when no topic is enriched.

The score is a linear function of the enrichment estimates, so its variance is a
quadratic form in their covariance $\Sigma$:

$$\mathrm{Var}(\hat c_i) = \sum_k \sum_m w_{ik}\, w_{im}\, \Sigma_{km}$$

summed over all ordered pairs. The p-value tests $c_i = 1$:

$$z_i = \frac{\hat c_i - 1}{\sqrt{\mathrm{Var}(\hat c_i)}}$$

## The cell-type heterogeneity test

For a cell type $g$, let $S_g$ be the covariance of the weight vectors $w$ across
its cells. The within-type variance of the score is

$$T_g = \mathrm{Var}_g(c) = \hat E' S_g \hat E$$

The null is that all topics are equally enriched, $E = c\mathbf{1}$ for an
unknown $c$ — not that nothing is enriched. If every topic shares the same
enrichment then every cell scores $c$ and the within-type variance is zero
regardless of the loadings.

Because the weights sum to one, $S_g\mathbf{1} = 0$ exactly, so writing
$\hat E = c\mathbf{1} + \varepsilon$ with $\varepsilon \sim N(0,\Sigma)$ removes
every term in $c$:

$$T_g = \varepsilon' S_g \varepsilon \sim \sum_j \lambda_j \chi^2_1, \qquad
\lambda = \mathrm{eig}\!\left(\Sigma^{1/2} S_g \Sigma^{1/2}\right)$$

The p-value is exact by Imhof inversion. No permutation is involved, and the
unknown level $c$ never has to be estimated.

## The covariance

Both quantities need $\Sigma$, the covariance of the topic enrichment estimates.
Obtain it with `ldsc_jackknife_cov()`, which reconstructs it from the
block-jackknife delete values S-LDSC writes under `--print-delete-vals`:

```{r sigma, eval=FALSE}
jk <- ldsc_jackknife_cov(
  ldsc_res_dir    = "path/to/scads_run",
  trait           = "IBD",
  nTopics         = 25,
  baseline_prefix = ".../1000G_Phase3_baselineLD_v2.2_ldscores/baselineLD.",
  frq_prefix      = ".../1000G_Phase3_frq/1000G.EUR.QC."
)

jk$pass         # TRUE if every topic reproduces S-LDSC's Enrichment and SE
head(jk$check)  # the per-topic comparison
```

`baseline_prefix` and `frq_prefix` must match the `--ref-ld-chr` and
`--frqfile-chr` of the runs being reconstructed. The acceptance check exists to
catch a mismatch: if the prefixes are wrong, `pass` is `FALSE`. A 25-topic run
takes a few minutes, and the result serves every trait sharing that topic model.

## Cell scores and p-values

Pass the reference prefixes to `get_cs()` and it builds $\Sigma$ for you:

```{r cs, eval=FALSE}
cs_res <- get_cs(
  topic_res       = topic_res,
  ldsc_res_dir    = "path/to/scads_run",
  trait           = "IBD",
  nTopics         = 25,
  baseline_prefix = ".../baselineLD.",
  frq_prefix      = ".../1000G.EUR.QC."
)

cs_res$cs        # per-cell score
cs_res$p_cell    # per-cell p-value, length = number of cells
cs_res$var_cell  # per-cell variance

fdr <- p.adjust(cs_res$p_cell, "BH")
```

`p_cell` is `NA` where the variance is zero, so it always aligns with `cs`.
`alternative = "greater"` gives a one-sided test of enrichment; the default is
two-sided, which also flags significantly depleted cells.

An already-computed covariance can be supplied instead with `Sigma = jk$Sigma`.

## Cell-type heterogeneity

```{r hetero, eval=FALSE}
res <- cell_type_heterogeneity(
  topic_res  = topic_res,
  enrichment = jk$enrichment,
  Sigma      = jk$Sigma,
  groups     = cell_type_labels   # one label per cell, in Lmat row order
)

subset(res, fdr < 0.05)
```

`groups` can be published cell-type labels, clusters, or any partition of the
cells. Groups smaller than `min_cells` (default 20) are skipped and `NA` labels
dropped. The result gives `cell_type`, `n`, the statistic, `p` and BH-adjusted
`fdr`, sorted by p-value.

## Interpreting the results

**Score spread does not predict heterogeneity.** $T_g$ and its null eigenvalues
both scale linearly with $S_g$, so the p-value is invariant to the overall size
of within-group variation. What matters is whether the direction in which cells
vary aligns with the direction in which topic enrichments differ. A cell type
with a wide score range can be non-significant, and a narrow one significant.

**Test association before heterogeneity.** Heterogeneity within a cell type that
carries no overall disease signal is heterogeneity in noise.

**Power follows the precision of the topic enrichments**, not the number of
cells. The test sees only differences between topics, so a trait with large but
imprecisely estimated enrichments can give no detections while a trait with a
smaller, tighter spread detects readily.

**Loadings are treated as fixed.** Topic-model uncertainty is not propagated into
either variance.

## Following up a significant cell type

The test says that cells within a cell type differ; it does not say which cells,
or along what axis. Correlating the cell score against the topic loadings within
that cell type identifies the topics driving the gradient, and their annotations
point at the regulatory programs responsible.
