---
title: "Testing cell types for heterogeneous disease relevance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing cell types for heterogeneous disease relevance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

SCADS gives every cell a disease-relevance score. Averaging those within a cell
type answers "is this cell type relevant?", but discards a different question:
do the cells *within* it differ from one another? `cell_type_heterogeneity()`
tests that, returning one p-value per cell type.

## Why the score makes this tractable

The cell score is a weighted average of the per-topic heritability enrichments,

$$s_i = w_i' E, \qquad \sum_k w_{ik} = 1$$

with weights determined by the cell's topic loadings and the topic annotation
sizes. For a cell type $g$, writing $S_g$ for the covariance of $w$ across its
cells, the within-type variance of the score is a quadratic form,

$$T_g = \hat E' S_g \hat E$$

The null of no heterogeneity is that every topic is equally enriched,
$E = c\mathbf{1}$ for some unknown $c$ — **not** that nothing is enriched. If all
topics share the same enrichment, every cell scores $c$ and the variance is zero
whatever the loadings.

Because the weights sum to one, $S_g\mathbf{1} = 0$ exactly, so $c$ cancels out
of the null entirely and

$$T_g \sim \sum_j \lambda_j \chi^2_1$$

with $\lambda$ the eigenvalues of $\Sigma^{1/2} S_g \Sigma^{1/2}$. The p-value is
exact by Imhof inversion — no permutation, no resampling.

That $c$ cancels is what makes the test usable: every topic annotation is open
chromatin, which is genuinely enriched for heritability, so $c \gg 1$ in
practice. A test that assumed $c = 1$ would reject in nearly every cell type,
detecting only that the annotations are peaks.

## You need an accurate covariance

The test needs $\Sigma$, the covariance of the topic enrichment estimates. SCADS
approximates this elsewhere as
$se_k\,se_m \mathrm{Cor}(A_k, A_m)$, using the correlation between annotations in
place of the correlation between their LD scores.

**That approximation is not adequate here.** It is fine for the cell score, whose
weight vector has a large component along $\mathbf{1}$, but this statistic is
built to be orthogonal to $\mathbf{1}$, which is exactly where the approximation
errs. On null simulations it makes the test conservative at every threshold,
whereas a block-jackknife $\Sigma$ is calibrated down to $p < 0.05$.

Use `ldsc_jackknife_cov()`. It requires your S-LDSC runs to have been made with
`--print-delete-vals`, and costs about three minutes for a 25-topic run.

```{r sigma, eval=FALSE}
jk <- ldsc_jackknife_cov(
  ldsc_res_dir    = "path/to/scads_run",
  trait           = "IBD",
  nTopics         = 25,
  baseline_prefix = ".../1000G_Phase3_baselineLD_v2.2_ldscores/baselineLD.",
  frq_prefix      = ".../1000G_Phase3_frq/1000G.EUR.QC."
)

jk$pass        # TRUE if every topic reproduces S-LDSC's own Enrichment and SE
head(jk$check) # the per-topic comparison
```

`baseline_prefix` and `frq_prefix` must match the `--ref-ld-chr` and
`--frqfile-chr` of the runs you are reconstructing; different runs in the same
project may use different references. The acceptance check exists to catch this:
if the prefixes are wrong, `pass` is `FALSE` and the reconstructed values will not
match the reported ones.

## Running the test

```{r test, eval=FALSE}
res <- cell_type_heterogeneity(
  topic_res  = topic_res,        # from run_fastTopics(): needs Pmat and Lmat
  enrichment = jk$enrichment,
  Sigma      = jk$Sigma,
  groups     = cell_type_labels  # one label per cell, in Lmat row order
)

subset(res, fdr < 0.05)
```

`groups` is whatever grouping you want tested — published cell-type labels,
clusters, or any partition of the cells. Groups smaller than `min_cells`
(default 20) are skipped, and `NA` labels are dropped.

The result has one row per cell type with the statistic, p-value and BH-adjusted
FDR, sorted by significance.

## Reading the output

**Score spread does not predict significance.** $T_g$ and its null eigenvalues
both scale linearly with $S_g$, so the p-value is invariant to the overall size
of within-group variation. What matters is whether the *direction* in which cells
vary aligns with the direction in which topic enrichments differ. A cell type with
a wide score range can be non-significant and a narrow one significant, so a
range-based summary and this test will disagree — by construction, not by error.

**Test association first.** Heterogeneity within a cell type that carries no
overall disease signal is heterogeneity in noise. Interpret these p-values only
for cell types that are associated to begin with.

**Power tracks the precision of the topic enrichments**, not the number of cells.
The test sees only *differences* between topics, so a trait whose enrichments are
large but imprecisely estimated can yield no detections at all while a trait with
a smaller but tighter spread detects readily.

**The loadings are treated as fixed.** Topic-model uncertainty is not propagated,
the same limitation that applies to the cell score variance.

## What the test does not do

It does not identify *which* cells drive the heterogeneity, or describe its
structure. Once a cell type is flagged, correlating the score against topic
loadings within that cell type is the natural follow-up, and will point at the
regulatory programs responsible.
