---
title: "Quick Tutorial on scads"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Prerequisites

Before running this tutorial, make sure you have completed all installation steps described in the [README](https://github.com/szhaolab/scads#installation):

1. **polyFUN** and **LDSC** are cloned and the `polyfun` conda environment is set up.
2. **scads** R package is installed with all dependencies.
3. **S-LDSC reference data** has been downloaded (see below).

## Download S-LDSC reference data

`scads` requires reference files from the S-LDSC baseline LD model. Download and extract them into a local directory (e.g., `~/LDSCORE/`). All files are available from [Zenodo](https://zenodo.org/records/10515792).

> **Disk usage:** The compressed downloads total ~1 GB. After extraction, the reference data requires ~4 GB of disk space.

```{bash eval=FALSE}
# Create a directory for the reference data
mkdir -p ~/LDSCORE

# Download reference files from Zenodo (~1 GB total)
cd ~/LDSCORE
curl -fSL -o 1000G_Phase3_plinkfiles.tgz \
  "https://zenodo.org/records/10515792/files/1000G_Phase3_plinkfiles.tgz?download=1"            # 275 MB
curl -fSL -o 1000G_Phase3_baselineLD_v2.2_ldscores.tgz \
  "https://zenodo.org/records/10515792/files/1000G_Phase3_baselineLD_v2.2_ldscores.tgz?download=1"  # 645 MB
curl -fSL -o 1000G_Phase3_frq.tgz \
  "https://zenodo.org/records/10515792/files/1000G_Phase3_frq.tgz?download=1"                   # 82 MB
curl -fSL -o 1000G_Phase3_weights_hm3_no_MHC.tgz \
  "https://zenodo.org/records/10515792/files/1000G_Phase3_weights_hm3_no_MHC.tgz?download=1"    # 12 MB
curl -fSL -o hm3_no_MHC.list.txt \
  "https://zenodo.org/records/10515792/files/hm3_no_MHC.list.txt?download=1"                    # 12 MB

# Extract archives
tar -xzf 1000G_Phase3_plinkfiles.tgz
tar -xzf 1000G_Phase3_frq.tgz
tar -xzf 1000G_Phase3_weights_hm3_no_MHC.tgz

# The baseline LD archive extracts files directly (not into a subdirectory),
# so create one first:
mkdir -p 1000G_Phase3_baselineLD_v2.2_ldscores
tar -xzf 1000G_Phase3_baselineLD_v2.2_ldscores.tgz -C 1000G_Phase3_baselineLD_v2.2_ldscores

# (Optional) Remove the .tgz files to free ~1 GB of disk space
# rm -f *.tgz
```

After extraction, you should have the following directory structure:

```
~/LDSCORE/
├── 1000G_EUR_Phase3_plink/          # 1000G.EUR.QC.{1..22}.{bed,bim,fam}
├── 1000G_Phase3_baselineLD_v2.2_ldscores/  # baselineLD.{1..22}.{annot.gz,l2.ldscore.gz,...}
├── 1000G_Phase3_frq/                # 1000G.EUR.QC.{1..22}.frq
├── 1000G_Phase3_weights_hm3_no_MHC/ # weights.hm3_noMHC.{1..22}.l2.ldscore.gz
└── hm3_no_MHC.list.txt
```

## Start an R session

Activate the `polyfun` conda environment **before** starting R. This ensures that `python3` (called internally by `scads`) resolves to the environment with the required Python packages.

```{bash eval=FALSE}
conda activate polyfun
R
```

Load the scads package:

```{r setup}
library(scads)
library(data.table)
```

## Load example data

`scads` ships with a small demo count matrix (peak-by-cell) and simulated GWAS summary statistics for chromosome 10.

```{r}
# Load demo peak-by-cell count matrix
data("demo_count_matrix")

# Examine the dataset
dim(demo_count_matrix)
demo_count_matrix[1:5, 1:5]
```

Load the example GWAS summary statistics:

```{r}
data_file <- system.file("extdata", "SIM_sumstats_chr10.txt.gz", package = "scads")
sumstats <- fread(data_file, header = TRUE)
head(sumstats)
```

## Configure paths

Set the paths to match your local environment. You need to specify:

- **`polyfun_d`** / **`ldsc_d`**: where polyFUN and LDSC are cloned.
- **`ldscore_dir`**: where the S-LDSC reference data was downloaded.
- **`output_dir`**: where scads will write its results.

```{r}
# ── User-specific paths (edit these) ──────────────────────────────────────────
polyfun_d    <- path.expand("~/polyfun/")
ldsc_d       <- path.expand("~/ldsc/")
ldscore_dir  <- path.expand("~/LDSCORE/")
output_dir   <- path.expand("~/scads_output/")

# ── Reference data paths (derived from ldscore_dir) ──────────────────────────
onekg_pref    <- file.path(ldscore_dir, "1000G_EUR_Phase3_plink", "1000G.EUR.QC.")
baseline_pref <- file.path(ldscore_dir, "1000G_Phase3_baselineLD_v2.2_ldscores", "baselineLD.")
frq_pref      <- file.path(ldscore_dir, "1000G_Phase3_frq", "1000G.EUR.QC.")
hm3_file      <- file.path(ldscore_dir, "hm3_no_MHC.list.txt")
weights_pref  <- file.path(ldscore_dir, "1000G_Phase3_weights_hm3_no_MHC", "weights.hm3_noMHC.")
```

Verify that the key files exist before proceeding:

```{r}
stopifnot(
  "polyfun ldsc.py not found"    = file.exists(file.path(polyfun_d, "ldsc.py")),
  "ldsc make_annot.py not found" = file.exists(file.path(ldsc_d, "make_annot.py")),
  "1000G plink files not found"  = file.exists(paste0(onekg_pref, "10.bim")),
  "Baseline LD files not found"  = file.exists(paste0(baseline_pref, "10.l2.ldscore.gz")),
  "Frequency files not found"    = file.exists(paste0(frq_pref, "10.frq")),
  "HapMap3 SNP list not found"   = file.exists(hm3_file),
  "Weight files not found"       = file.exists(paste0(weights_pref, "10.l2.ldscore.gz"))
)
cat("All paths verified.\n")
```

## Set analysis parameters

```{r}
count_matrix <- demo_count_matrix
n_topics     <- 5
n_sims       <- 1000
n_cores      <- 8        # adjust to the number of cores available on your machine
bl_method    <- "gc"
bl_ct        <- NULL
bl_peaks     <- NULL
sumstats_d   <- dirname(data_file)
gwas_n       <- 764667
trait         <- "SIM"
cont_topics  <- FALSE
```

## Run scads

The `scads()` function runs the full pipeline:

1. **fastTopics** — fits a topic model on the count matrix.
2. **Create topic annotations** — generates BED files for each topic.
3. **S-LDSC** — runs stratified LD score regression for each topic annotation using the baseline LD v2.2 model.
4. **Calculate cell scores** — computes a per-cell disease score.

> **Note:** This tutorial runs on **chromosome 10 only** (`chrs = 10`) to keep runtime short. For a real analysis, use the default `chrs = 1:22` to run on all autosomes.

```{r}
set.seed(1234)
scads_res <- scads(
  count_matrix,
  nTopics                = n_topics,
  n_s                    = n_sims,
  n_c                    = n_cores,
  baseline_method        = bl_method,
  bl_celltype            = bl_ct,
  bl_celltype_peak_file  = bl_peaks,
  sumstats_dir           = sumstats_d,
  gwas_nsamps            = gwas_n,
  gwas_trait             = trait,
  outdir                 = output_dir,
  polyfun_code_dir       = polyfun_d,
  ldsc_code_dir          = ldsc_d,
  onekg_path             = onekg_pref,
  baseline_dir           = baseline_pref,
  frqfile_pref           = frq_pref,
  hm3_snps               = hm3_file,
  weights_pref           = weights_pref,
  continuous_topic_annot = cont_topics,
  chrs                   = 10,
  genome                 = "hg19"
)
```

## Examine results

```{r}
# Save results
saveRDS(scads_res, file = file.path(output_dir, "scads_res.rds"))

# Cell scores
summary(scads_res$cs_res$cs)
```

The result is a list with two elements:

- **`out1`**: the fitted fastTopics model and topic annotations.
- **`cs_res`**: the per-cell disease scores.
