---
title: "Quick Tutorial on scads"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Tutorial on scads}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup 

Confirm [PolyFUN](https://github.com/omerwe/polyfun) and [LDSC](https://github.com/bulik/ldsc) exist and runs in your conda environment. See their websites for installation and usage guidance.

```{bash eval=F}
# srun --mem-per-cpu=8GB --cpus-per-task=16 --time=8:00:00 --pty /bin/bash
conda activate /dartfs/rc/lab/S/Szhao/liyang/conda/base_clone/
# Rscript -e "rmarkdown::render('/dartfs/rc/lab/S/Szhao/liyang/scads/vignettes/Introduction.Rmd', output_file='/dartfs/rc/lab/S/Szhao/liyang/scads/vignettes/Introduction.html', output_format='html_document')"
R
```

Load R packages

```{r setup}
# library(remotes)
# remotes::install_github("szhaolab/scads")
# library(scads)

setwd("/dartfs/rc/lab/S/Szhao/liyang/scads")
devtools::load_all()
library(scads)
```

Load example data

```{r}
# Access the dataset
data("demo_count_matrix")

# Examine dataset
dim(demo_count_matrix)
demo_count_matrix[1:5,1:5]
```

Prepare a GWAS summary statistics file 

```{r}
data_file <- system.file("extdata", "SIM_sumstats_chr10.txt.gz", package = "scads")
sumstats <- fread(data_file, header = T)
head(sumstats)
```

## Run scads

1. Run fastTopics 
  
2. Create topic annotations 

3. Run S-LDSC using topic annotations

  - We will run hg19 version of S-LDSC using the recommended baseline_v2.2 model 
  
4. Calculate Cell Score 


```{r}
count_matrix = demo_count_matrix
n_topics = 5
n_sims = 1000
n_cores = 16
bl_method = "gc" 
bl_ct = NULL
bl_peaks =NULL
sumstats_d = dirname(data_file)
gwas_n = 764667
trait = "SIM"

myHomeDir <- "/dartfs/rc/lab/S/Szhao/"
# /dartfs/rc/lab/S/Szhao/liyang/SCSviaTM/analysis/

output_dir = paste0(myHomeDir, "liyang/SCSviaTM/analysis/test_260318_gc/")
polyfun_d= paste0(myHomeDir, "liyang/polyfun/")
ldsc_d = paste0(myHomeDir, "liyang/ldsc/")
onekg_pref = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/1000G_EUR_Phase3_plink/1000G.EUR.QC.")
# onekg_pref = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/1000G_EUR_Phase3_plink/1000G.EUR.QC.10.")

baseline_pref = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/1000G_Phase3_baselineLD_v2.2_ldscores/baselineLD.")
# baseline_pref = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/1000G_Phase3_baselineLD_v2.2_ldscores/baselineLD.10.")

frq_pref = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/1000G_Phase3_frq/1000G.EUR.QC.")
# frq_pref = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/1000G_Phase3_frq/1000G.EUR.QC.10.")

hm3_file = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/hm3_no_MHC.list.txt")
weights_pref = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/1000G_Phase3_weights_hm3_no_MHC/weights.hm3_noMHC.")
# weights_pref = paste0(myHomeDir, "fine-mapping/LDSCORE/zenodo/1000G_Phase3_weights_hm3_no_MHC/weights.hm3_noMHC.10.")

cont_topics = FALSE
```

```{r}
set.seed(1234)
scads_res <- scads(
  count_matrix,
  nTopics                 = n_topics,
  n_s                     = n_sims,
  n_c                     = n_cores,
  baseline_method         = bl_method,
  bl_celltype             = bl_ct,
  bl_celltype_peak_file   = bl_peaks,
  sumstats_dir            = sumstats_d,
  gwas_nsamps             = gwas_n,
  gwas_trait              = trait,
  outdir                  = output_dir,
  polyfun_code_dir        = polyfun_d,
  ldsc_code_dir           = ldsc_d,
  onekg_path              = onekg_pref,
  baseline_dir            = baseline_pref,
  frqfile_pref            = frq_pref,
  hm3_snps                = hm3_file,
  weights_pref            = weights_pref,
  continuous_topic_annot  = cont_topics, 
  chrs                    = 10,
  genome                  = "hg19"
)

saveRDS(scads_res, file = paste0(output_dir, "scads_res.rds"))
```
  


