---
title: "Quick Tutorial on scads"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Tutorial on scads}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup 

Confirm [PolyFUN](https://github.com/omerwe/polyfun) runs in your conda environment

```{bash eval=F}
conda activate /dartfs/rc/lab/S/Szhao/liyang/conda/base_clone/
```

Load R packages

```{r setup}
library(scads)
library(data.table)
```

Load example data

```{r}
# Access the dataset
data("demo_count_matrix")

# Examine dataset
dim(demo_count_matrix)
demo_count_matrix[1:5,1:5]
```

Prepare a GWAS summary statistics file 

```{r}
data_file <- system.file("extdata", "SIM_sumstats_chr10.txt.gz", package = "scads")
sumstats <- fread(data_file, header = T)
head(sumstats)
```

## Run scads

1. Run fastTopics 

  - fastTopics: fit model and obtain topics 
  - fastTopics-DE: obtain peaks associated with topics
  - locfdr: adjust for multiple testing [**issue with NA** on some runs - TODO: FIX ISUE]
    (Current resolution: just run again)
  
2. Create topic annotations for LDSC 

3. Run LDSC 

  - Resolved issues - should be running though the enrichment results will not make sense 
  
4. Calculate Cell Score 

  - Made up topic enrichment estimates (topic 2 is enriched)
  - Output list of 2
    1. out1 (fastTopics results)
    2. cs (cell scores)
  

```{r}
set.seed(1)
# using f0 = 1/(3E9/500)*100
scads_res <- scads(demo_count_matrix, 
                     nTopics = 5, 
                     bed_dir = "/dartfs/rc/lab/S/Szhao/liyang/SCSviaTM/analysis/simu_20241028/bed_dir", 
                     sumstats_dir = "/dartfs/rc/lab/S/Szhao/liyang/SCSviaTM/analysis/simu_20241028", 
                     gwas_nsamps = 45000, 
                     gwas_trait = "SIM", 
                     outdir = "/dartfs/rc/lab/S/Szhao/liyang/SCSviaTM/analysis/simu_20241028/scads")
```

