---
title: "Quick Start"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What is scads?

`scads` computes per-cell disease relevance scores from scATAC-seq data and
GWAS summary statistics. Give it a peak-by-cell count matrix and GWAS sumstats,
and it returns a score for every cell indicating how disease-relevant that
cell's chromatin profile is.

## Load example data

```{r setup}
library(scads)
library(ggplot2)
```

`scads` ships with a demo dataset: 500 cells, 12,971 peaks on chromosome 10,
and simulated GWAS summary statistics.

```{r}
data("demo_count_matrix")
cat("Count matrix:", nrow(demo_count_matrix), "peaks x",
    ncol(demo_count_matrix), "cells\n")
```

## Run the full pipeline

The `scads()` function runs the entire pipeline in one call. Before running,
you need to set up paths to [polyFUN](https://github.com/omerwe/polyfun),
[LDSC](https://github.com/bulik/ldsc), and the S-LDSC reference data
(see the [Step-by-Step Tutorial](StepByStep.html) for details on downloading
reference data).

```{r eval=FALSE}
data_file <- system.file("extdata", "SIM_sumstats_chr10.txt.gz", package = "scads")

set.seed(1234)
scads_res <- scads(
  demo_count_matrix,
  nTopics             = 5,
  n_s                 = 1000,
  n_c                 = 8,
  baseline_method     = "gc",
  sumstats_dir        = dirname(data_file),
  gwas_nsamps         = 45087,
  gwas_trait          = "SIM",
  outdir              = "~/scads_output/",
  polyfun_code_dir    = "~/polyfun/",
  ldsc_code_dir       = "~/ldsc/",
  onekg_path          = "~/LDSCORE/1000G_EUR_Phase3_plink/1000G.EUR.QC.",
  baseline_dir        = "~/LDSCORE/1000G_Phase3_baselineLD_v2.2_ldscores/baselineLD.",
  frqfile_pref        = "~/LDSCORE/1000G_Phase3_frq/1000G.EUR.QC.",
  hm3_snps            = "~/LDSCORE/hm3_no_MHC.list.txt",
  weights_pref        = "~/LDSCORE/1000G_Phase3_weights_hm3_no_MHC/weights.hm3_noMHC.",
  chrs                = 10,
  genome              = "hg19",
  save_intermediates  = TRUE,
  verbose             = TRUE
)
```

The pipeline takes ~50 minutes on a MacBook Pro (M3 Max) for this demo
(chr10 only, 5 topics). For the rest of this vignette, we load pre-computed
results so you can explore the output immediately.

## Explore results

```{r}
# Load pre-computed demo results
demo_res <- readRDS(system.file("extdata", "demo_results.rds", package = "scads"))
```

### S-LDSC enrichment per topic

Each topic represents a set of co-accessible peaks. S-LDSC tests whether
GWAS heritability is enriched in each topic's peaks. Enrichment > 1 means
the topic's peaks harbour more heritability than expected by chance.

```{r fig.width=6, fig.height=4}
ldsc_tab <- demo_res$ldsc_res_table
ldsc_tab$ASH_Enrichment <- demo_res$ash_PosteriorMean

ggplot(ldsc_tab, aes(x = Category, y = Enrichment)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_errorbar(aes(ymin = Enrichment - 1.96 * Enrichment_std_error,
                     ymax = Enrichment + 1.96 * Enrichment_std_error),
                width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(title = "S-LDSC enrichment per topic",
       subtitle = "Dashed line = no enrichment (1.0). Error bars = 95% CI.",
       x = "Topic", y = "Enrichment") +
  theme_minimal()
```

```{r}
print(ldsc_tab[, c("Category", "Prop._SNPs", "Prop._h2", "Enrichment",
                    "Enrichment_std_error", "ASH_Enrichment")])
```

### Cell scores

The final output is a per-cell disease score that combines topic enrichments
with each cell's topic loadings.

```{r}
cat("Cell score summary:\n")
print(summary(demo_res$cs))

cat("\nSignificant cells (p < 0.05):", sum(demo_res$p_cell < 0.05), "of",
    length(demo_res$p_cell), "\n")
```

### UMAP visualization

UMAP embeds cells in 2D based on their topic loadings. Coloring by cell score
reveals which cell populations are most disease-relevant.

```{r fig.width=10, fig.height=4, message=FALSE}
if (!requireNamespace("uwot", quietly = TRUE)) {
  message("Install 'uwot' for UMAP: install.packages('uwot')")
} else {
  library(patchwork)

  set.seed(42)
  umap_coords <- uwot::umap(demo_res$Lmat, n_neighbors = 350, min_dist = 0.8)

  umap_df <- data.frame(
    UMAP1 = umap_coords[, 1],
    UMAP2 = umap_coords[, 2],
    cell_score = demo_res$cs,
    z_score = as.numeric(demo_res$z_cell),
    topic = paste0("k", apply(demo_res$Lmat, 1, which.max))
  )

  p1 <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = topic)) +
    geom_point(size = 0.8, alpha = 0.7) +
    labs(title = "Dominant topic", color = "Topic") +
    theme_minimal() + theme(legend.position = "bottom")

  p2 <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cell_score)) +
    geom_point(size = 0.8, alpha = 0.7) +
    scale_color_viridis_c(option = "C") +
    labs(title = "Cell score", color = "Score") +
    theme_minimal() + theme(legend.position = "bottom")

  p3 <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = z_score)) +
    geom_point(size = 0.8, alpha = 0.7) +
    scale_color_gradient2(low = "blue", mid = "grey90", high = "red", midpoint = 0) +
    labs(title = "Z-score", color = "Z") +
    theme_minimal() + theme(legend.position = "bottom")

  p1 + p2 + p3 + plot_layout(ncol = 3)
}
```

Cells dominated by topics with high enrichment (e.g., k3, k4) show elevated
scores, while cells in low-enrichment topics score near the baseline.

## Next steps

- See the [Step-by-Step Tutorial](StepByStep.html) for a detailed walkthrough
  of each pipeline stage, diagnostic plots, and parameter explanations.
- For a real analysis, use `chrs = 1:22` to run on all chromosomes.
- Use `resume_from_step` to re-run specific steps without starting over.
