---
title: "Step-by-Step Tutorial"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Step-by-Step Tutorial}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`scads` (**S**ingle-**C**ell **A**TAC-seq **D**isease **S**coring) computes
per-cell disease relevance scores by combining single-cell chromatin
accessibility data with GWAS summary statistics. The pipeline has four main
steps:

| Step | What it does | Typical runtime | Peak memory |
|------|-------------|----------------|-------------|
| 1. **fastTopics** | Fits a topic model on the count matrix, identifies topic-specific peaks via differential expression, and computes a GC-content baseline correction. | ~1 min (500 cells, 13K peaks, 5 topics) | ~650 MB |
| 2. **Topic annotations** | Converts significant peaks per topic into BED files for S-LDSC. | < 10 sec | Negligible |
| 3. **S-LDSC** | Runs stratified LD score regression for each topic using the baseline LD v2.2 model. This is the most time-consuming step. | ~10 min per topic per chromosome (sequential) | ~300 MB per topic (Python) |
| 4. **Cell scores** | Combines S-LDSC enrichment estimates with cell-level topic loadings to produce a per-cell disease score. | < 10 sec | ~650 MB |

> **Note on runtime:** The times above were measured on a MacBook Pro (M3 Max,
> 36 GB RAM) using the demo dataset (500 cells, 12,971 peaks on chr10, 5
> topics). A full-genome analysis (22 chromosomes) will take proportionally
> longer for Step 3. Larger datasets (more cells or peaks) will increase
> Step 1 runtime.

## Prerequisites

Before running this tutorial, make sure you have completed all installation
steps described in the
[README](https://github.com/szhaolab/scads#installation):

1. **polyFUN** and **LDSC** are cloned and the `polyfun` conda environment is
   set up (Python 3.11, with `bitarray<3` and `bedtools` installed).
2. **scads** R package is installed with all dependencies.
3. **S-LDSC reference data** has been downloaded (see below).

## Download S-LDSC reference data

`scads` requires reference files from the S-LDSC baseline LD model. Download
and extract them into a local directory (e.g., `~/LDSCORE/`). All files are
available from [Zenodo](https://zenodo.org/records/10515792).

> **Disk usage:** The compressed downloads total ~1 GB. After extraction, the
> reference data requires ~4 GB of disk space.

```{bash eval=FALSE}
# Create a directory for the reference data
mkdir -p ~/LDSCORE

# Download reference files from Zenodo (~1 GB total)
cd ~/LDSCORE
curl -fSL -o 1000G_Phase3_plinkfiles.tgz \
  "https://zenodo.org/records/10515792/files/1000G_Phase3_plinkfiles.tgz?download=1"            # 275 MB
curl -fSL -o 1000G_Phase3_baselineLD_v2.2_ldscores.tgz \
  "https://zenodo.org/records/10515792/files/1000G_Phase3_baselineLD_v2.2_ldscores.tgz?download=1"  # 645 MB
curl -fSL -o 1000G_Phase3_frq.tgz \
  "https://zenodo.org/records/10515792/files/1000G_Phase3_frq.tgz?download=1"                   # 82 MB
curl -fSL -o 1000G_Phase3_weights_hm3_no_MHC.tgz \
  "https://zenodo.org/records/10515792/files/1000G_Phase3_weights_hm3_no_MHC.tgz?download=1"    # 12 MB
curl -fSL -o hm3_no_MHC.list.txt \
  "https://zenodo.org/records/10515792/files/hm3_no_MHC.list.txt?download=1"                    # 12 MB

# Extract archives
tar -xzf 1000G_Phase3_plinkfiles.tgz
tar -xzf 1000G_Phase3_frq.tgz
tar -xzf 1000G_Phase3_weights_hm3_no_MHC.tgz

# The baseline LD archive extracts files directly (not into a subdirectory),
# so create one first:
mkdir -p 1000G_Phase3_baselineLD_v2.2_ldscores
tar -xzf 1000G_Phase3_baselineLD_v2.2_ldscores.tgz -C 1000G_Phase3_baselineLD_v2.2_ldscores

# (Optional) Remove the .tgz files to free ~1 GB of disk space
# rm -f *.tgz
```

After extraction, you should have the following directory structure:

```
~/LDSCORE/
├── 1000G_EUR_Phase3_plink/          # 1000G.EUR.QC.{1..22}.{bed,bim,fam}
├── 1000G_Phase3_baselineLD_v2.2_ldscores/  # baselineLD.{1..22}.{annot.gz,l2.ldscore.gz,...}
├── 1000G_Phase3_frq/                # 1000G.EUR.QC.{1..22}.frq
├── 1000G_Phase3_weights_hm3_no_MHC/ # weights.hm3_noMHC.{1..22}.l2.ldscore.gz
└── hm3_no_MHC.list.txt
```

## Start an R session

Activate the `polyfun` conda environment **before** starting R. This ensures
that `python3` (called internally by `scads`) resolves to the environment with
the required Python packages (including `ldsc`, `polyfun`, `bedtools`,
`bitarray<3`).

```{bash eval=FALSE}
conda activate polyfun
R
```

Load the scads package:

```{r setup}
library(scads)
library(data.table)
library(ggplot2)
```


## Load example data

`scads` ships with a small demo dataset:

- **`demo_count_matrix`**: A sparse peak-by-cell count matrix (12,971 peaks on
  chr10 x 500 cells) from scATAC-seq data.
- **`SIM_sumstats_chr10.txt.gz`**: Simulated GWAS summary statistics for
  chromosome 10.

```{r}
# Load demo peak-by-cell count matrix
data("demo_count_matrix")

# Examine the dataset
cat("Dimensions (peaks x cells):", dim(demo_count_matrix), "\n")
cat("Matrix class:", class(demo_count_matrix), "\n")
cat("Non-zero entries:", length(demo_count_matrix@x), "\n")
cat("Sparsity:", round(1 - length(demo_count_matrix@x) / prod(dim(demo_count_matrix)), 4), "\n")
demo_count_matrix[1:5, 1:5]
```

The count matrix is a sparse matrix where rows are genomic peaks (in
`chr:start-end` format) and columns are cells. Most entries are zero, which is
typical for scATAC-seq data.

Load the example GWAS summary statistics:

```{r}
data_file <- system.file("extdata", "SIM_sumstats_chr10.txt.gz", package = "scads")
sumstats <- fread(data_file, header = TRUE)
cat("Summary statistics dimensions:", dim(sumstats), "\n")
cat("Columns:", paste(names(sumstats), collapse = ", "), "\n")
head(sumstats)
```

The summary statistics file must contain columns: `chr`, `pos`, `beta`, `se`,
`a0` (ref allele), `a1` (alt allele), `snp` (rsID), `pval`, and `zscore`.

## Configure paths

Set the paths to match your local environment. You need to specify:

- **`polyfun_d`** / **`ldsc_d`**: where polyFUN and LDSC are cloned.
- **`ldscore_dir`**: where the S-LDSC reference data was downloaded.
- **`output_dir`**: where scads will write its results.

```{r}
# -- User-specific paths (edit these) -----------------------------------------
polyfun_d    <- path.expand("~/polyfun/")
ldsc_d       <- path.expand("~/ldsc/")
ldscore_dir  <- path.expand("~/Desktop/Project_scads/LDSCORE/zenodo/")
output_dir   <- path.expand("~/scads_output/")

# -- Reference data paths (derived from ldscore_dir) --------------------------
onekg_pref    <- file.path(ldscore_dir, "1000G_EUR_Phase3_plink", "1000G.EUR.QC.")
baseline_pref <- file.path(ldscore_dir, "1000G_Phase3_baselineLD_v2.2_ldscores", "baselineLD.")
frq_pref      <- file.path(ldscore_dir, "1000G_Phase3_frq", "1000G.EUR.QC.")
hm3_file      <- file.path(ldscore_dir, "hm3_no_MHC.list.txt")
weights_pref  <- file.path(ldscore_dir, "1000G_Phase3_weights_hm3_no_MHC", "weights.hm3_noMHC.")
```

Verify that the key files exist before proceeding:

```{r}
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
stopifnot(
  "polyfun ldsc.py not found"    = file.exists(file.path(polyfun_d, "ldsc.py")),
  "ldsc make_annot.py not found" = file.exists(file.path(ldsc_d, "make_annot.py")),
  "1000G plink files not found"  = file.exists(paste0(onekg_pref, "10.bim")),
  "Baseline LD files not found"  = file.exists(paste0(baseline_pref, "10.l2.ldscore.gz")),
  "Frequency files not found"    = file.exists(paste0(frq_pref, "10.frq")),
  "HapMap3 SNP list not found"   = file.exists(hm3_file),
  "Weight files not found"       = file.exists(paste0(weights_pref, "10.l2.ldscore.gz"))
)
cat("All paths verified.\n")
```

## Set analysis parameters

```{r}
count_matrix <- demo_count_matrix
n_topics     <- 5       # number of topics to fit
n_sims       <- 1000    # Monte Carlo samples for DE analysis
n_cores      <- 8       # adjust to the number of cores available
bl_method    <- "gc"    # GC-content baseline correction (recommended)
bl_ct        <- NULL    # cell type for "estimate" baseline (not used here)
bl_peaks     <- NULL    # cell type peaks for "estimate" baseline (not used here)
sumstats_d   <- dirname(data_file)
gwas_n       <- 45087   # GWAS sample size
trait         <- "SIM"
cont_topics  <- FALSE   # use binary (not continuous) topic annotations
```

### Baseline method options

The `baseline_method` parameter controls how the null hypothesis is set for
differential expression testing in Step 1:

| Method | Description |
|--------|-------------|
| `"gc"` | **(Recommended)** Fits a GC-content-dependent baseline using an EM algorithm with a Poisson mixture model. Accounts for GC bias in chromatin accessibility. |
| `"constant"` | Uses a fixed small constant (`1e-7`) as baseline. Simple but does not account for GC bias. |
| `"average"` | Computes a genome-wide average accessibility rate. |
| `"estimate"` | Estimates background from a reference cell type's non-overlapping peaks. Requires `bl_celltype` and `bl_celltype_peak_file`. |

---

## Running the pipeline step by step

While `scads()` can run the entire pipeline in one call, this tutorial walks
through each step individually to show intermediate results. The `scads()`
wrapper calls these same functions internally and supports
**checkpointing** via `save_intermediates = TRUE` and `resume_from_step`.

> **Note:** This tutorial runs on **chromosome 10 only** (`chrs = 10`) to keep
> runtime short (~50 min total). For a real analysis, use `chrs = 1:22`.

### Step 1: Fit topic model with fastTopics

**What this step does:**

1. Transposes the count matrix (peaks x cells -> cells x peaks) for fastTopics.
2. Fits a Poisson NMF topic model with `k` topics using `fastTopics::fit_topic_model()`.
3. Computes a GC-content baseline correction for each topic (if `baseline_method = "gc"`).
4. Runs differential expression analysis to identify topic-specific peaks.
5. Applies FDR correction and binarizes peak assignments.

**Runtime:** ~1 min for the demo dataset (500 cells, 12,971 peaks, 5 topics).
**Memory:** ~650 MB peak.

```{r}
out1_cache <- file.path(output_dir, "run_fastTopics_res.rds")
if (file.exists(out1_cache)) {
  cat("Loading cached Step 1 results from", out1_cache, "\n")
  out1 <- readRDS(out1_cache)
} else {
  set.seed(1234)

  # Transpose: scads expects peaks-by-cells, but fastTopics needs cells-by-peaks
  count_matrix_t <- Matrix::t(count_matrix)
  cat("Input dimensions (cells x peaks):", dim(count_matrix_t), "\n")

  # Run fastTopics
  t1_start <- Sys.time()
  out1 <- run_fastTopics(
    count_matrix  = count_matrix_t,
    nTopics       = n_topics,
    n_s           = n_sims,
    n_c           = n_cores,
    baseline_method = bl_method,
    bl_celltype_cells = bl_ct,
    bl_celltype_peak_file = bl_peaks,
    fdr_cutoff    = 0.05,
    outdir        = output_dir,
    genome        = "hg19"
  )
  t1_end <- Sys.time()
  cat("\nStep 1 runtime:", round(difftime(t1_end, t1_start, units = "secs")), "seconds\n")

  # Save intermediate results
  saveRDS(out1, out1_cache)
}
```

**Examine Step 1 results:**

```{r}
# Topic model dimensions
cat("Factor matrix F (peaks x topics):", dim(out1$Fmat), "\n")
cat("Loading matrix L (cells x topics):", dim(out1$Lmat), "\n")
cat("Binary peak matrix P (peaks x topics):", dim(out1$Pmat), "\n")

# Number of significant peaks per topic (after FDR correction)
cat("\nSignificant peaks per topic:\n")
print(colSums(out1$Pmat))
```

The binary peak matrix `Pmat` indicates which peaks are significantly enriched
in each topic (1 = significant at FDR < 0.05). These will be used as topic
annotations for S-LDSC.

### Step 2: Generate topic annotation BED files

**What this step does:**

Converts the binary peak-topic assignments (`Pmat`) into BED files for each
topic. Each BED file contains the genomic coordinates of peaks assigned to that
topic, which S-LDSC uses as a custom annotation.

**Runtime:** < 10 seconds.
**Memory:** Negligible.

```{r}
beddir_cache <- file.path(output_dir, "beddir_list.rds")
if (file.exists(beddir_cache)) {
  cat("Loading cached Step 2 results from", beddir_cache, "\n")
  beddir_list <- readRDS(beddir_cache)
} else {
  t2_start <- Sys.time()
  beddir_list <- get_annot_beds(topics_res = out1, output_dir = output_dir)
  t2_end <- Sys.time()
  cat("Step 2 runtime:", round(difftime(t2_end, t2_start, units = "secs"), 1), "seconds\n")

  # Save for checkpointing
  saveRDS(beddir_list, beddir_cache)
}
```

### Step 3: Run S-LDSC per topic

**What this step does:**

For each topic annotation, this step runs the full S-LDSC pipeline:

1. **Munge summary statistics** -- converts GWAS sumstats to the format expected
   by LDSC (parquet file via `munge_polyfun_sumstats.py`).
2. **Create annotation** -- for each chromosome, marks which 1000G SNPs fall
   within topic peaks (via `make_annot.py`).
3. **Compute LD scores** -- calculates LD scores for the custom annotation
   (via `ldsc.py --l2`).
4. **Heritability estimation** -- runs partitioned heritability analysis
   combining the custom annotation with the baseline LD v2.2 model
   (via `ldsc.py --h2`).

The key output is the **enrichment estimate** per topic: how much heritability
is concentrated in the topic's peaks relative to their size.

**Runtime:** ~10 min per topic per chromosome. With 5 topics on chr10 running
sequentially, expect ~50 min total.
**Memory:** ~300 MB per topic (Python processes).

```{r}
# Check which topics already have S-LDSC results
t3_start <- Sys.time()
all_done <- all(vapply(seq_along(beddir_list), function(i) {
  file.exists(file.path(beddir_list[[i]], "results", paste0(trait, ".results")))
}, logical(1)))

if (all_done) {
  cat("All S-LDSC results found — skipping Step 3.\n")
} else {
  for (i in seq_along(beddir_list)) {
    res_file <- file.path(beddir_list[[i]], "results", paste0(trait, ".results"))
    if (file.exists(res_file)) {
      cat(sprintf("--- Topic %d already done, skipping ---\n", i))
      next
    }
    cat(sprintf("\n--- Running S-LDSC for topic %d of %d ---\n", i, length(beddir_list)))

    run_sldsc(
      chrs          = 10,                    # chr10 only for this tutorial
      polyfun_path  = polyfun_d,
      ldsc_path     = ldsc_d,
      sumstats_path = sumstats_d,
      n             = gwas_n,
      trait         = trait,
      onekg_path    = onekg_pref,
      bed_dir       = beddir_list[[i]],
      baseline_dir  = baseline_pref,
      frqfile_pref  = frq_pref,
      hm3_snps      = hm3_file,
      weights_pref  = weights_pref,
      out_dir       = beddir_list[[i]]
    )
  }
}
t3_end <- Sys.time()
cat("\nStep 3 total runtime:", round(difftime(t3_end, t3_start, units = "secs"), 1), "seconds\n")
```

The enrichment results per topic are displayed after Step 4 (which reads and
processes the S-LDSC output files).

### Step 4: Calculate cell scores

**What this step does:**

Combines the S-LDSC enrichment estimates with cell-level topic loadings to
produce a per-cell disease relevance score:

1. Reads S-LDSC results for each topic and applies adaptive shrinkage (ASH) to
   stabilize enrichment estimates.
2. Computes a weighted score per cell: `cs_i = sum(L_ik * a_k * e_k) / sum(L_ik * a_k)`,
   where `L_ik` is the cell's loading on topic `k`, `a_k` is the number of
   peaks in topic `k`, and `e_k` is the (shrunk) enrichment estimate.
3. Computes z-scores and p-values accounting for correlation between topic
   annotations.

**Runtime:** < 10 seconds.
**Memory:** ~650 MB (due to correlation matrix computation).

```{r}
cs_cache <- file.path(output_dir, "cs_res.rds")
if (file.exists(cs_cache)) {
  cat("Loading cached Step 4 results from", cs_cache, "\n")
  cs_res <- readRDS(cs_cache)
} else {
  t4_start <- Sys.time()
  cs_res <- get_cs(
    topic_res    = out1,
    ldsc_res_dir = output_dir,
    trait        = trait,
    nTopics      = n_topics
  )
  t4_end <- Sys.time()
  cat("Step 4 runtime:", round(difftime(t4_end, t4_start, units = "secs"), 1), "seconds\n")

  # Save results
  saveRDS(cs_res, cs_cache)
}
```

**Examine Step 4 results:**

```{r}
cat("Cell score summary:\n")
print(summary(cs_res$cs))

cat("\nZ-score summary:\n")
print(summary(as.numeric(cs_res$z_cell)))

cat("\nNumber of significant cells (p < 0.05):", sum(cs_res$p_cell < 0.05), "\n")
cat("Number of significant cells (p < 0.01):", sum(cs_res$p_cell < 0.01), "\n")

cat("\nASH-shrunk enrichment estimates:\n")
print(cs_res$ash_res$PosteriorMean)
```

**Understanding the output:**

- **`cs`**: The per-cell disease score. Higher values indicate greater disease
  relevance based on the cell's chromatin accessibility profile.
- **`z_cell`**: Z-score for each cell, accounting for variance in enrichment
  estimates and correlation between topic annotations.
- **`p_cell`**: Two-sided p-value for each cell's score.
- **`ldsc_res_table`**: Raw S-LDSC results per topic.
- **`ash_res`**: Enrichment estimates after adaptive shrinkage (more stable).
- **`cs_dat`**: Intermediate values (`M_i`, `N_i`) used in score computation.

Higher cell scores indicate cells whose chromatin accessibility profiles overlap
more strongly with disease-associated regions. Cells dominated by topics with
high enrichment will receive higher scores.

---

## Visualizing results

After running the pipeline, it is important to inspect intermediate and final
results. This section shows diagnostic plots for each stage.

### Topic model diagnostics

#### Distribution of factor loadings (Fmat)

The factor loading matrix `Fmat` (peaks x topics) contains the Poisson rate
parameters for each peak in each topic. Examining the distribution of log10(F)
per topic reveals whether the model separates foreground (high F) from
background (low F) peaks.

```{r fig.width=8, fig.height=5}
Fmat <- out1$Fmat

# Histogram of log10(Fmat) per topic
fmat_df <- do.call(rbind, lapply(1:ncol(Fmat), function(k) {
  data.frame(
    log10F = log10(Fmat[, k]),
    topic  = paste0("k", k)
  )
}))

ggplot(fmat_df, aes(x = log10F)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white", linewidth = 0.1) +
  facet_wrap(~topic, scales = "free_y") +
  labs(title = "Distribution of log10(F) per topic",
       x = "log10(F)", y = "Count") +
  theme_minimal()
```

#### Structure plot

The `fastTopics` structure plot shows how cells are distributed across topics.
Each vertical bar represents a cell, colored by its topic proportions (from the
`Lmat` loading matrix). Cells dominated by a single topic form clusters.

```{r fig.width=8, fig.height=3}
fastTopics::structure_plot(out1$fastTopics_fit)
```

### S-LDSC enrichment summary

#### Enrichment bar plot

Visualizing enrichment estimates with error bars helps identify which topics
carry disease signal. Topics with positive enrichment (and small standard error)
are the informative ones.

```{r fig.width=6, fig.height=4}
ldsc_tab <- cs_res$ldsc_res_table

# Add ASH-shrunk estimates
ldsc_tab$ASH_Enrichment <- cs_res$ash_res$PosteriorMean

# Bar plot of enrichment per topic
ggplot(ldsc_tab, aes(x = Category, y = Enrichment)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_errorbar(aes(ymin = Enrichment - 1.96 * Enrichment_std_error,
                     ymax = Enrichment + 1.96 * Enrichment_std_error),
                width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(title = "S-LDSC enrichment per topic",
       subtitle = "Dashed line = no enrichment (1.0). Error bars = 95% CI.",
       x = "Topic", y = "Enrichment") +
  theme_minimal()
```

#### S-LDSC results table

```{r}
# Full results table
print(ldsc_tab[, c("Category", "Prop._SNPs", "Prop._h2", "Enrichment",
                    "Enrichment_std_error", "ASH_Enrichment")])
```

> **Interpreting enrichment:** Enrichment > 1 means heritability is
> concentrated in a topic's peaks more than expected by chance. Topics with
> enrichment significantly > 1 (confidence interval not crossing 1) indicate
> disease-relevant chromatin programs. After adaptive shrinkage (ASH), the
> estimates are stabilized and used to weight cell scores in Step 4.

### Cell score distribution

#### Histogram of cell scores

```{r fig.width=6, fig.height=4}
cs_df <- data.frame(cs = cs_res$cs)

ggplot(cs_df, aes(x = cs)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Distribution of cell scores",
       x = "Cell score", y = "Count") +
  theme_minimal()
```

#### Z-score distribution

The z-scores account for variance in enrichment estimates and correlation
between topic annotations. Under the null (no disease signal), z-scores should
be approximately standard normal.

```{r fig.width=6, fig.height=4}
z_df <- data.frame(z = as.numeric(cs_res$z_cell))
z_df <- z_df[is.finite(z_df$z), , drop = FALSE]

ggplot(z_df, aes(x = z)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(-1.96, 1.96), linetype = "dashed", color = "red") +
  labs(title = "Distribution of cell z-scores",
       subtitle = "Dashed lines = +/-1.96 (p = 0.05 threshold)",
       x = "Z-score", y = "Count") +
  theme_minimal()
```

### UMAP visualization

UMAP provides a 2D embedding of cells based on the topic loading matrix (`Lmat`).
Coloring cells by their disease score reveals whether disease-relevant cells
cluster together.

```{r fig.width=10, fig.height=4, message=FALSE}
if (!requireNamespace("uwot", quietly = TRUE)) {
  message("Install 'uwot' for UMAP: install.packages('uwot')")
} else {
  library(patchwork)

  # Compute UMAP from topic loadings
  set.seed(42)
  umap_coords <- uwot::umap(out1$Lmat, n_neighbors = 350, min_dist = 0.8)

  umap_df <- data.frame(
    UMAP1 = umap_coords[, 1],
    UMAP2 = umap_coords[, 2],
    cell_score = cs_res$cs,
    z_score = as.numeric(cs_res$z_cell)
  )

  # Color by dominant topic
  dominant_topic <- paste0("k", apply(out1$Lmat, 1, which.max))
  umap_df$topic <- dominant_topic

  p3 <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = topic)) +
    geom_point(size = 0.8, alpha = 0.7) +
    labs(title = "Dominant topic", color = "Topic") +
    theme_minimal() + theme(legend.position = "bottom")

  # Color by cell score
  p1 <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cell_score)) +
    geom_point(size = 0.8, alpha = 0.7) +
    scale_color_viridis_c(option = "C") +
    labs(title = "Cell score", color = "Score") +
    theme_minimal() + theme(legend.position = "bottom")

  # Color by z-score
  p2 <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = z_score)) +
    geom_point(size = 0.8, alpha = 0.7) +
    scale_color_gradient2(low = "blue", mid = "grey90", high = "red", midpoint = 0) +
    labs(title = "Z-score", color = "Z") +
    theme_minimal() + theme(legend.position = "bottom")

  # Combine plots
  p3 + p1 + p2 + plot_layout(ncol = 3)
}
```

> **Note:** Cells dominated by topics with high enrichment will appear as
> "hot spots" on the score UMAP. The z-score UMAP highlights cells whose
> scores are statistically significant after accounting for estimation
> uncertainty and topic correlation.

---

## Output files

After running the pipeline, `output_dir` contains:

```
output_dir/
├── run_fastTopics_res.rds          # Step 1 results (fastTopics model + DE)
├── fasTopics_fit.rds               # Raw fastTopics model fit
├── count_mat_tp.rds                # Pseudobulk count matrix (peaks x topics)
├── beddir_list.rds                 # Step 2 results (BED directory paths)
├── cs_res.rds                      # Step 4 results (cell scores)
├── k1_output/                      # Topic 1
│   ├── k1_annotations.bed          #   BED file of topic peaks
│   ├── annotations/SIM/            #   S-LDSC annotation files
│   │   └── SIM.10.{annot.gz,l2.ldscore.gz,...}
│   ├── results/                    #   S-LDSC results
│   │   └── SIM.{results,log}
│   └── SIM_munged_sumstats.parquet #   Munged GWAS sumstats
├── k2_output/                      # Topic 2 (same structure)
│   └── ...
└── k5_output/                      # Topic 5
    └── ...
```

The result list returned by `scads()` has two elements:

- **`out1`**: The fitted fastTopics model, including:
  - `fastTopics_fit`: The raw topic model object
  - `Fmat`: Factor loadings matrix (peaks x topics)
  - `Lmat`: Topic proportions matrix (cells x topics)
  - `Pmat`: Binary peak assignments (peaks x topics)
  - `de_res`: Full differential expression results (z-scores, log-fold changes)
  - `baseline`: The GC-content baseline used for DE testing

- **`cs_res`**: The per-cell disease scores, including:
  - `cs`: Cell score vector
  - `z_cell`: Z-scores per cell
  - `p_cell`: P-values per cell
  - `ldsc_res_table`: S-LDSC results per topic
  - `ash_res`: ASH-stabilized enrichment estimates
  - `cs_dat`: Intermediate `M_i` and `N_i` values

## Troubleshooting

### Common issues

| Issue | Solution |
|-------|---------|
| `"sortBed" does not appear to be installed` | Install bedtools in your conda env: `conda install -n polyfun -c bioconda bedtools` |
| `TypeError: float() argument must be a string or a real number, not 'bitarray.decodeiterator'` | Downgrade bitarray: `pip install 'bitarray<3'` |
| `could not find function "width"` | Make sure `GenomicRanges` is loaded (scads handles this internally) |
| Paths with spaces break S-LDSC | scads v0.2.0+ handles this via `shQuote()`. If using older versions, use symlinks. |
| S-LDSC takes too long | Use `chrs = 10` for testing. For production, run on a cluster with `mc.cores`. |
